Protein significance analysis of mass spectrometry-based proteomics experiments with R and MSstats (v3.0.9)

### 2015-09-30 MSstats_manual_RMarkdown.Rmd
Sections 3-4

MSstats is an open-source R-based package for statistical relative quantification of peptides and proteins in mass spectrometry-based proteomic experiments. This document describes `MSstats`, the most recent version of the package, and its use through the command line.

@@

# 3. Prerequisites

Set the working directory to where you saved files. Note that you should a different path on your computer from the example.

@@

If you didn't have MSstats installed so far, you can downlaod the package from the [MSstats installation page](http://msstats.org/installation/) to your working directory and install it as follows:

Load MSstats into the R session and verify that you have the correct version (3.0.9) loaded.

@@

Controlled mixture DDA data will be used for demonstration. This dataset is available in [MSstats material github](https://github.com/MeenaChoi/MSstats_material) in the folder named 'example dataset/DDA_controlledMixture2009'. It is processed by Superhirn. [original reference link](http://prottools.ethz.ch/muellelu/web/Latin_Square_Data.php)

@@

Let's start by reading in data.

@@

@@

(**! Always pay attention to the default options **)

The default option for normalization is `equalizeMedians`, where all the intensities in a run are shifted by a constant, to equalize the median of intensities across runs for label-free experiment. This normalization works when the assumption that the majority of proteins do not change across runs is held. Therefore if your label-free DDA dataset has a small number of proteins, please be careful to use the `equalizeMedians` option. Instead, if you have a spiked in standard then you want to set this to `globalStandards` and define the standard with `nameStandards`. For label based experiment, `equalizeMedians` equalizes the median of reference intensities across runs for label based experiment. Therefore, it works for dataset with a small number of proteins.

Usually label-free DDA datasets have many missing values and noisy features with outliers. MSstats supports several ways to deal with this. The default option for summarization is `TMP` (Tukey's median polish, robust parameter estimation method with median across rows and columns) after imputation by AFT (accelerated failure time model, `MBimpute=TRUE`) based on censored intensity for NA (`censoredInt="NA"`) with a cutoff as the minimum value for a feature and a run (`cutoffCensored="minFeatureNRun"`). Here the NA values represent censored measurements, which are below the cutoff and not detected. This process handles missing values through imputation and reduces the influence of the outliers using the TMP estimation. However, those runs with no measurements at all will be removed and not be used for any calculation.

```{r, eval=T}
# default option
DDA2009.TMP <- dataProcess(raw = DDA2009.superhirn,  fillIncompleteRows = TRUE,
                           normalization = 'equalizeMedians',
                           summaryMethod = 'TMP',
                           censoredInt = "NA", cutoffCensored="minFeatureNRun",
                           MBimpute = TRUE)
```

If `MBimpute=FALSE`, imputation will not be performed and censored intensities (here NAs) will be just replaced with cutoff value assigned in `cutoffCensored="minFeatureNRun"`. With this option, those runs with substantial missing measurements will be biased by the cutoff value. In such case, you can remove the runs that have more than 50% missing values from your analysis with `remove50missing=TRUE`.

@@

If `censoredInt=NULL`, we assume that all intensities are randomly missing and there is no action for missing values.

```{r, eval=T}
# no action for missing values.
DDA2009.TMP.random <- dataProcess(raw = DDA2009.superhirn, fillIncompleteRows = TRUE,

                           normalization = 'equalizeMedians',
                           summaryMethod = 'TMP',
                           censoredInt=NULL)
```

@@

Besides summarizing observations with the median polish method, MSstats also offers a summarization option using linear model, and one based on the sum of log-intensities, which is the default Skyline behaviour. Summarization using the option `summaryMethod="linear"` with `censoredInt=NULL` assumes that all NAs are randomly missing and uses `lm` or `lmer` for parameter estimation. On the other hand, `summaryMethod="linear"` with `censoredInt="NA"` assumes that NA intensities are censored, not detected because intensities are below the cutoff. AFT model is then applied with left-censored for parameter estimation. 'cutoffCensored' is the same as `summaryMethod="TMP"`. For `summaryMethod="linear"`, AFT model is only used for parameter estimation, not for imputation. Therefore `MBimpute=TRUE` with `summaryMethod="linear"` doesn't work (no imputation before parameter estimation, the same with `MBimpute=FALSE`).

@@

#### 4.2.2.1 Quality control and Normalization effects

Now let's look at what the equalize medians procedure did to our data. We can generate QC plots for all proteins or for single proteins at a time if we have a big dataset. Let's look at the 'Kininogen-1' (Kng1/NP_036828) protein in these example data.

@@

***NOTE***
Don't worry about warning messages as below. It means they will not plot with NA values, which we want.

Warning messages:
1: In loop_apply(n, do.ply) :
  Removed 698 rows containing non-finite values (stat_boxplot).

@@

Let's look at how the original data look like. Profile plot shows all individual measurements per protein. With `type="profileplot"`, two pdfs will be generated. The first pdf includes plots (per protein) to show individual measurement for each peptide (peptide for DDA, transition for SRM or DIA) across runs, grouped per condition. Each peptide has a different color/type layout. If you don't want to generate these plots, please use the option `originalPlot=FALSE`. The second pdf, which is named with 'wSummarization' suffix, shows run-level summarized data per protein. The same peptides (or transition) in the first plot are presented in grey, with the summarized values by TMP overlaid in red.

@@

Have a look at the profile plots to compare the different options (e.g., TMP with or without considering missing values for summarization in `dataProcess`.) Original profile plots are the same. But, summarization plots are different, especially for conditions 'C1' and 'C2' in 'yeast' protein, which have many missing values.

@@

After we normalized the data and summarized each protein's behaviour across conditions with one of the `dataProcess` summarization methods, we are all set to compare protein changes between groups of conditions. Within MSstats we can do this with the `groupComparison` function, which takes as input the output of the `dataProcess` function.

@@

We need to provide the `groupComparison` function with a contrast matrix to define the comparison to be made. The contrast matrix is created with each condition in column and each comparison in row. Note that the conditions are arranged **in alphabetical order**:

@@

It shows the order of condition, which MSstats can recognize. Then, we add a row with entries of 0, 1, or -1 for every comparison we would like to make between groups of conditions.

**0** is for conditions we would like to ignore.
**1** is for conditions we would like to put in the numerator of the ratio or fold-change.
**-1** is for conditions we would like to put in the denumerator of the ratio or fold-change.

For example, if you want to compare C2-C1, which means log(C2)-log(C1) and the same as log(C2/C1), set '1' for C2 and '-1' for C1 in the row. Also, combining multiple groups for comparison is possible. For example, if you want to compare between average of C2 and C3 and average of C1, (C3+C2)/2-C1 as formula, set '-1' for C1, '0.5' for C2 and '0.5' for C3, and '0' for rest of groups.

@@

We're ready to go! let's compare our two populations.

```{r, eval=T, message=F, warning=F}
DDA2009.comparisons <- groupComparison(contrast.matrix = comparison, data=DDA2009.TMP)
```

Let's inspect the results to see what proteins change significantly between groups.

@@

You can get a shorter list of the differentially abundant proteins based on the comparison results:

```{r, eval=TRUE}
# get only significant proteins and comparisons among all comparisons
SignificantProteins = 
  DDA2009.comparisons$ComparisonResult[DDA2009.comparisons$ComparisonResult$adj.pvalue < 0.05 ,]
nrow(SignificantProteins)
```

@@

Volcano plots visualize the outcome of one comparison between conditions for all the proteins, and combine the information on statistical and practical significance. The y-axis displays
the FDR-adjusted p-values on the negative log10 scale, and represents statistical significance.
The horizontal dashed line represents the FDR cutoff. The points above the FDR cutoff line
are statistically significant differentially abundant proteins. These points are colored in red for
upregulated proteins, and in blue for downregulated proteins. The x-axis is the model-based
estimate of log-fold change (the base of logarithm transform is the same as specified in the
logTrans option of the dataProcess step), and represents practical significance. It is possible
to specify a practical significance cutoff based on the estimate of fold change in addition to
the statistical significance cutoff. If the fold change cutoff is specified, the points above the
horizontal cutoff line but within the vertical cutoff line will be judged as not differentially
abundant (and will be colored in black). The practical significance cutoff can only be applied
in addition to the statistical significance cutoff (i.e. the fold change alone does not present
enough evidence for differential abundance).

@@

Then, 'VolcanoPlot.pdf' will be saved under the folder you assigned. It has the plots per comparison you set in contrast.matrix. Below is one of volcano plot, for comparison 'C6-C5'. Please check `?groupComparisonPlots` for detail, such as labelling protein names, size of dots, font sizes, etc.

@@

Heatmaps illustrate the patterns of up- and down-regulation of proteins in several comparisons. Columns in the heatmaps are comparison of conditions you assigned in `contrast.matrix`, and rows are proteins. The heatmaps display signed FDR-adjusted p-values of the tests, colored in red/blue for significantly up-/down-regulated proteins, while taking into account the specified FDR cutoff and the additional optional fold change cutoff. Brighter colors
indicate stronger evidence in favor of differential abundance. Black color represents proteins
are not significantly differentially abundant.

@@

Then, 'VolcanoPlot.pdf' will be saved under the folder you assigned. It has the plots per comparison you set in contrast.matrix. Below is one of volcano plot, for comparison 'C6-C5'. Please check `?groupComparisonPlots` for detail, such as labelling protein names, size of dots, font sizes, etc.

>>>>

`MSstats` is an open-source R-based package for statistical relative quantification of peptides and proteins in mass spectrometry-based proteomic experiments. This document describes `MSstats`, the most recent version of the package, and its use through the command line.

@@

# 3. Prerequisites and setting for MSstats analysis

`MSstats` is an R-based package. It is assumed that you already have R installed. You can downlaod the package from the [MSstats installation page](http://msstats.org/installation/) and install it as follows:

Alternatively, you can install `MSstats` from Bioconductor:

```{r, eval=F}
source("https://bioconductor.org/biocLite.R")
biocLite("MSstats")
```

Once you have the package installed, load `MSstats` into an R session and verify that you have the correct version (3.0.9). Note that in order to use `MSstats`, the package needs to be loaded every time you restart R.

Finally, set the working directory to where you saved files. Note that you may have a different path on your computer from the example.

@@

@@

This section describes a typical workflow for DDA analysis with `MSstats`. Controlled mixture DDA data will be used for demonstration. This dataset (RawData.DDA.csv) is available in [MSstats material github](https://github.com/MeenaChoi/MSstats_material) in the folder named 'example dataset/DDA_controlledMixture2009'. It is processed by Superhirn. [original reference link](http://prottools.ethz.ch/muellelu/web/Latin_Square_Data.php)

@@

The first step in using the `MSstats` is to format the data as described in Section~2. We load here the prepared example dataset and confirm the data is stored in the correct format.

@@

To get started with this function, visit the help section of `dataProcess` first:

@@

`dataProcess` provides a variety of options in consideration of different experimental protocols. Therefore, the default options may not be appropriate for all possible scenarios. It is important to understand their underlying assumption to avoid misuse.

The default option for normalization is `equalizeMedians`, where all the intensities in a run are shifted by a constant, to equalize the median of intensities across runs for label-free experiment. This normalization method is appropriate when the majority of proteins do not change across runs. Be cautious when using the `equalizeMedians` option for a label-free DDA dataset with only a small number of proteins. Instead, if you have a spiked in standard, you may set this to `globalStandards` and define the standard with `nameStandards`. For label based experiment, `equalizeMedians` equalizes the median of reference intensities across runs and is generally proper even for a dataset with a small number of proteins.

A typical label-free DDA dataset may have many missing values and noisy features with outliers. `MSstats` supports several ways to deal with this. The default option for summarization is `TMP` (Tukey's median polish, robust parameter estimation method with median across rows and columns) after imputation by AFT (accelerated failure time model, `MBimpute=TRUE`) based on censored intensity for `NA` (`censoredInt="NA"`) with a cutoff as the minimum value for a feature and a run (`cutoffCensored="minFeatureNRun"`). Here the `NA` values represent censored measurements, which are below the cutoff and undetectable. This process handles missing values through imputation and reduces the influence of the outliers using the `TMP` estimation. Note, however, that those runs with no measurements at all will be removed and not be used for any calculation.

```{r, eval=T}
# default option
DDA2009.TMP <- dataProcess(raw = DDA2009.superhirn,  fillIncompleteRows = TRUE,
                           normalization = 'equalizeMedians',
                           summaryMethod = 'TMP',
                           censoredInt = "NA", cutoffCensored = "minFeatureNRun",
                           MBimpute = TRUE)
```

If `MBimpute=FALSE`, imputation will not be performed and censored intensities (here `NA`'s) will be replaced with a cutoff value specified in `cutoffCensored`. The default option for `cutoffCensored` is `minFeatureNRun`, taking the minimum value for the corresponding feature and run. With this option, those runs with substantial missing measurements will be biased by the cutoff value. In such case, you may remove the runs that have more than 50% missing values from the analysis with the option `remove50missing=TRUE`.

Output of the `dataProcess` function contains the processed and run-level summarized data as well as relevant information for the summarization step.

@@

If `censoredInt=NULL`, we assume that all intensities are missing at random and there is no action for missing values.

```{r, eval=T}
# no action for missing values.
DDA2009.TMP.random <- dataProcess(raw = DDA2009.superhirn, fillIncompleteRows = TRUE,
                                  normalization = 'equalizeMedians',
                                  summaryMethod = 'TMP',
                                  censoredInt=NULL)
```

@@

Besides summarizing observations with the median polish method, `MSstats` also offers a summarization option using linear model, and one based on the sum of log-intensities, which is the default Skyline operation Summarization using the option `summaryMethod="linear"` with `censoredInt=NULL` assumes that all `NA`'s are missing at random and uses `lm` or `lmer` for parameter estimation. On the other hand, `summaryMethod="linear"` with `censoredInt="NA"` assumes that `NA` intensities are censored, not detected because the intensities are below the cutoff. AFT model is then applied with left-censored for parameter estimation. `cutoffCensored` is the same as `summaryMethod="TMP"`. For `summaryMethod="linear"`, AFT model is only used for parameter estimation, not for imputation. Therefore using `summaryMethod="linear"` with `MBimpute=TRUE` will yield the same result as with `MBimpute=FALSE` (no imputation before parameter estimation).

@@

#### 4.2.2.1 Quality control and normalization effects

QC plots are used to assess the effects of the normalization step. These plots can be generated for either all proteins at once or each protein individually if we have a large dataset. The example below shows both options.

@@

***NOTE***
Don't worry about warning messages as below. It means `NA` values are not included in the plot, which is a proper way for this case.

```
Warning messages:
1: In loop_apply(n, do.ply) :
  Removed 698 rows containing non-finite values (stat_boxplot).
```

@@

Profile plots present the effects of the summarization step by showing all individual measurements of a protein and their summarized intensity. With `type="profileplot"`, two pdfs will be generated. The first pdf includes plots (per protein) to show individual measurement for each peptide (peptide for DDA, transition for SRM or DIA) across runs, grouped per condition. Each peptide has a different color/type layout. To ignore these plots, please use the option `originalPlot=FALSE`. The second pdf, which is named with 'wSummarization' suffix, shows run-level summarized data per protein. The same peptides (or transition) in the first plot are presented in grey, with the summarized values (by TMP, in this example) overlaid in red.

@@

These plots can be used compare and select among different options for summarization (e.g., `TMP` with or without considering missing values for summarization in `dataProcess`). While original profile plots are the same, summarization plots reveal differences, especially for conditions 'C1' and 'C2' in 'yeast' protein, which have many missing values.

@@

With the normalized data and run-level summarized data obtained by applying one of the `dataProcess` summarization methods, it is of general interest to find proteins changing between groups of conditions. Within `MSstats` this can be done by using the `groupComparison` function, which takes as input the output of the `dataProcess` function.

@@

In addition to the processed data, the `groupComparison` function requires a contrast matrix to define the comparison to be made. The contrast matrix is created with each condition in column and each comparison in row. Note that the conditions are arranged **in alphabetical order**. The order of condition that `MSstats` recognizes can be shown by using `levels`:

@@

Entries in each row of the contrast matrix are filled in with 0, 1, or -1 to specify the comparison, where **0** is for conditions we would like to ignore, **1** is for conditions we would like to put in the numerator of the ratio or fold-change, and **-1** is for conditions we would like to put in the denumerator of the ratio or fold-change.

For example, if you want to compare C2-C1, which means log(C2)-log(C1) and the same as log(C2/C1), set '1' for C2 and '-1' for C1 in the row. Combining multiple groups for comparison is also possible. For example, if you want to compare between average of C2 and C3 and average of C1, (C3+C2)/2-C1 as formula, set '-1' for C1, '0.5' for C2 and '0.5' for C3, and '0' for rest of groups.

@@

With the contrast matrix specified, group comparison can be performed as follows.

```{r, eval=T, message=F, warning=F}
DDA2009.comparisons <- groupComparison(contrast.matrix = comparison, data = DDA2009.TMP)
```

Output of the `groupComparison` function contains three data frames:

@@

Based on the comparison results and desired significance level, a short list of the differentially abundant proteins can be obtained for further investigation:

```{r, eval=TRUE}
# get only significant proteins and comparisons among all comparisons
SignificantProteins <- with(DDA2009.comparisons, ComparisonResult[ComparisonResult$adj.pvalue < 0.05, ])
nrow(SignificantProteins)
```

@@

Volcano plots visualize the outcome of one comparison between conditions for all the proteins, and combine the information on statistical and practical significance. The y-axis displays the FDR-adjusted p-values on the negative log10 scale, representing statistical significance. The horizontal dashed line shows the FDR cutoff. The points above the FDR cutoff line are statistically significant proteins that are differentially abundant across conditions. These points are colored in red and blue for upregulated and downregulated proteins, respectively. The x-axis is the model-based estimate of fold change on log scale (the base of logarithm transform is the same as specified in the `logTrans` option of the `dataProcess` function), and represents practical significance. It is possible to specify a practical significance cutoff based on the estimate of fold change in addition to the statistical significance cutoff. If the fold change cutoff is specified, the points above the horizontal cutoff line but within the vertical cutoff line will be considered as not differentially abundant (and will be colored in black). The practical significance cutoff should only be applied in addition to the statistical significance cutoff (i.e., the fold change alone does not present enough evidence for differential abundance).

@@

'VolcanoPlot.pdf' will be saved under the folder you assigned. It has the plots per comparison defined in `contrast.matrix`. Below is one of volcano plots, for comparison 'C6-C5'. Please check `?groupComparisonPlots` for detail, such as labelling protein names, size of dots, font sizes, etc.

@@

Heatmaps illustrate the patterns of up- and down-regulation of proteins in several comparisons. Columns in the heatmaps are comparison of conditions assigned in `contrast.matrix`, and rows are proteins. The heatmaps display signed FDR-adjusted p-values of the tests, colored in red/blue for significantly up-/down-regulated proteins, while taking into account the specified FDR cutoff and the additional optional fold change cutoff. Brighter colors indicate stronger evidence in favor of differential abundance. Black color represents proteins that are not significantly differentially abundant.

@@

'Heatmap.pdf' will be saved under the folder you assigned. It has the plots per comparison defined in `contrast.matrix`. Below is one example, showing the results for several comparisons simultaneously. For further details, such as labelling protein names, size of dots, font sizes, etc., please check `?groupComparisonPlots`.
